---
title: "OutTakes.Rmd"
author: "Jim Callahan"
date: "October 25, 2015"
output: pdf_document
---

paper describing the experiment suggested that there were "96 features" (variables).

> "a wearable sensor-oriented classification approach for the detection of mistakes"  

> "mounted the sensors in the users' glove, armband, lumbar belt and
> dumbbell (see Figure 1)."

**Sensors  (from Figure 1)**  

- ArmBand (on body)  
- Belt (on body)  
- Glove (on body)  
- Dumbbell (on weight)  

"For data recording we used four 9 degrees of freedom Razor
inertial measurement units (IMU), 
which provide three-axes acceleration, gyroscope and magnetometer data at a joint
sampling rate of 45 Hz."

**Class**  
> "**Class A** corresponds to the [correct] specified execution of the exercise,  
> while the other 4 classes correspond to common mistakes."  
  
A. Correct -- "exactly according to the specification"  
B. Incorrect -- "throwing the elbows to the front"  
C. Incorrect -- "lifting the dumbbell only halfway"  
D. Incorrect -- "lowering the dumbbell only halfway"  
E. Incorrect -- "throwing the hips to the front"  
  
  

**Window**  


**Euler Angles**  
> "Euler angles (roll, pitch and yaw)"  

-Roll  
-Pitch  
-Yaw  

**Calculated 8 Features**  
> "For the **[three] Euler angles** of each of the **four sensors**  
> we **calculated eight features**:  
> mean, variance, standard deviation, max, min, amplitude, kurtosis and skewness,  
> generating in total **96 derived feature sets.**"  

3 Euler angles  
4 Sensors  
8 Calculated Features  

```{r}
3*4*8
```




**Quality**
> "if we can specify how an activity has to be performed we can measure the quality  
> by comparing its execution against this specification.  
>  
> From this, we define quality as the adherence of the execution of an activity  
> to its specification.  
>  
> From this, we define a qualitative activity recognition system as a   
> software artefact that observes the user's execution of an activity and
> compares it to a specification."


**Drop "X" the ID number to prevent "data leakage"**  
The variable "**X**" appears to be a sequence number and will have to be discarded prior
to training to avoid "data leakage".  

```{r}
summary(rawtraining$X)
```


The sequence number should be non-informative (useless) but if it turns out to be predictive it would be an example of "**data leakage**".  
  
  
Kaggle defines **data leakage** as, "the creation of unexpected additional information 
in the training data, allowing a model or machine learning algorithm to make unrealistically 
good predictions."  
https://www.kaggle.com/wiki/Leakage

An similar example of data leakage would be:  
"You’re trying to study who has breast cancer.  The **patient ID**, which seemed 
innocent, actually has predictive power. What happened?  ...This is probably a 
consequence of using multiple databases [each from different cancer centers], some 
of which correspond to [specialize in] sicker patients are more likely to be sick."  
This blog post corresponds to pages 310-311 in the book, "**Doing Data Science**".  
http://mathbabe.org/2012/11/20/columbia-data-science-course-week-12-predictive-modeling-data-leakage-model-evaluation/



> "For the **[three] Euler angles** of each of the **four sensors**  we  calculated 
> **eight features**:   
> mean, variance, standard deviation, max, min, amplitude, kurtosis and skewness,  
> generating in total **96 derived feature sets.**"  
    
**3 Euler angles**  
  - roll, pitch and yaw    
**4 Sensors**  
  - belt, arm ("armband"), dumbell and forearm ("glove")  
**8 Calculated Features**  
  - mean, variance, standard deviation, max, min, amplitude, kurtosis and skewness     
  
```{r}
3*4*8
```
